---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# {ao}

<!-- badges: start -->
[![R-CMD-check](https://github.com/loelschlaeger/ao/workflows/R-CMD-check/badge.svg)](https://github.com/loelschlaeger/ao/actions)
[![CRAN status](https://www.r-pkg.org/badges/version-last-release/ao)](https://www.r-pkg.org/badges/version-last-release/ao)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/grand-total/ao)](https://cranlogs.r-pkg.org/badges/grand-total/ao)
[![Codecov test coverage](https://codecov.io/gh/loelschlaeger/ao/branch/main/graph/badge.svg)](https://app.codecov.io/gh/loelschlaeger/ao?branch=main)
<!-- badges: end -->

This R package implements alternating optimization, which is an iterative procedure for optimizing some function jointly over all parameters by alternating restricted optimization over individual parameter subsets. 

For more details see the help vignette:
``` r
vignette("ao", package = "ao")
```

## Installation

You can install the released version from [CRAN](https://CRAN.R-project.org) with:

```{r, install released, eval = FALSE}
install.packages("ao")
```

And the development version from [GitHub](https://github.com/) with:

```{r, install dev, eval = FALSE}
# install.packages("devtools")
devtools::install_github("loelschlaeger/ao")
```

## Example

The following lines perform alternating optimization of the [Himmelblau's function](https://en.wikipedia.org/wiki/Himmelblau%27s_function), separately for $x_1$ and $x_2$, with the parameter restrictions $-5 \leq x_1, x_2 \leq 5$:

```{r, example}
library("ao")
himmelblau <- function(x) (x[1]^2 + x[2] - 11)^2 + (x[1] + x[2]^2 - 7)^2
ao(
  f = himmelblau, p = c(0,0), partition = list(1, 2),
  optimizer = set_optimizer_optim(lower = -5, upper = 5, method = "L-BFGS-B")
)
```
