---
title: "ao"
author: "Lennart Oelschläger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ao}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ao)
```

## Theory

### What is alternating optimization?

informal definition

### Where is alternating optimization applied?

names + references

When is alternating optimization a good choice? page 8

### How is the alternating optimization algorithm defined?

formal definition

ingredients

explicit vs implicit 

two options: complete alternating or partly alternating

### What are the properties of alternating optimization?

existence and uniqueness of optima

convergence speed

convergence rate

## Applications

Example 1

```{r example 1}
f <- set_f(f = function(x) x[1]^2 - 3*x[1]*x[2] + x[2]^2, npar = 2)
ao(f = f, initial = -1, partition = list(1,2))
ao(f = f, initial = 1, partition = list(c(1,2)))
```

Example 2

```{r example 2}
A <- matrix(c(100, 80, 5, 1, 80, 90, 2, 1, 5, 2, 70, 40, 1, 1, 40, 80), 4, 4)
q <- function(x) as.numeric(t(x) %*% A %*% x)
f <- set_f(f = q, npar = 4)
ao(f = f, initial = 1, partition = list(1:2,3:4), iterations = 100)
ao(f = f, initial = 1, partition = list(1:4), iterations = 100)
```

Example from Mail

```{r example 3}
valley <- function(x) {
  cons <- c(1.003344481605351, -3.344481605351171e-03)
  n <- length(x)
  f <- rep(0, n)
  j <- 3 * (1:(n/3))
  jm2 <- j - 2
  jm1 <- j - 1
  f[jm2] <- (cons[2]*x[jm2]^3 + cons[1]*x[jm2]) * exp(-(x[jm2]^2)/100) - 1
  f[jm1] <- 10 * (sin(x[jm2]) - x[jm1])
  f[j] <- 10 * (cos(x[jm2]) - x[j])
  sum(f*f)
}
f <- set_f(f = valley, npar = 9, lower = 0, upper = 10, check = FALSE)
ao(f = f, partition = list(1, 2, 3, 4, 5, 6, 7, 8, 9), initial = 0, iterations = 3)
```
